---
title: "Analysis of US ETFs"
author: "Nabil Jabareen"
date: "31 März 2018"
output: 
  html_document: 
    code_folding: hide
---

```{r warning=FALSE, include=FALSE}
setwd("/Users/Anja/Desktop/Kaggle_StockMarket/")
#install_github("walkerke/bsselectR")

library(data.table)
library(ggplot2)
library(magrittr) #imports "%>%" option
library(gridExtra) #multi grid s for ggplots
library(plotly) #interactie plot
library(psych) #correlation/pairs plot
library(corrplot) #correlation plot
```

#The Data  
Given are 1344 ETFs from the US. As shown below, the data consists of the Variables Date, Open, High, Low, Close, Volume and OpenInt. The Variable *Price* has been added for every Date by taking the mean of the *High* and *Low* for every date.

```{r Read ETFs, cache=TRUE, echo=FALSE}
#Read ETF files in a list of data.tables
etfs.files <- list.files("ETFs", pattern = "*.txt")

etfs <- lapply(etfs.files,
               function(x) fread(paste("ETFs" ,x, sep = "/"), header = T, sep = ","))

names(etfs) <- gsub(".txt", "", etfs.files)


#Convert dates to date Formats
etfs <- lapply(etfs, function(x) {
  x$Date <- as.Date(x$Date)
  x
})

#Calculate mean for every date
etfs <- lapply(etfs, function(x) {
  x[, Price := apply(x, 1, function(y) mean(as.numeric(c(y["High"], y["Low"])) ))]
  })

#Structure of ETF data
head(etfs[[1]])

#Get data.table with date and Price only
etfs.red <- lapply(etfs, function(x) x[,c("Date", "Price")])
```

```{r overall return}

overall_return <- function(x){
  o.return <- (x[which.max(Date)]$Price - x[which.min(Date)]$Price)/x[which.min(Date)]$Price
  o.return
  }

overall_returns <- sapply(etfs.red, function(x) overall_return(x))
```

With the price given for each ETF, I calculated the [fractal dimension](https://en.wikipedia.org/wiki/Fractal_dimension) first introduced by Benoit Mandelbrot. The fractal dimension can be thought of as a measure of roughness for a given geometric object (including curves). To calculate the fractal dimension for finencial price charts various methods have been suggested. Some of them can get quite complicated. I choosed a rather simple one by John Ehlers ( [Original publication](http://www.stockspotter.com/Files/frama.pdf)). The Formula to calculate the fractal dimension $D$ is as follows:

$$\begin{aligned}
D &= \frac{Log(HL1 + HL2) - Log(HL)}{Log(2)} \\
HL1 &= \frac{Max(High, \frac{1}{2}N..N)-Min(Low,\frac{1}{2}N..N)}{\frac{1}{2}N} \\
HL2 &= \frac{Max(High, \frac{1}{2}N)-Min(Low,\frac{1}{2}N)}{\frac{1}{2}N} \\
HL &= \frac{(Max(High,N) - Min(Low,N))}{N}
\end{aligned}$$

I applied this formula on all ETFs with 1000 or more data points. 

```{r Fractal Dimensions, warning=FALSE, cache=TRUE}
#Remove etfs with very low data
etf.length <- sapply(etfs.red, function(x) nrow(x))
etfs.red <- etfs.red[etf.length > 999]

#Function to calculate fractal dimension as intruduced by John Ehlers
fractal_dimension <- function(data){
  
  data.half <- nrow(data)/2
  first.half <- 1:round(data.half)
  data1 <- data[first.half]
  data2 <- data[!first.half]
  
  hl1 <- (max(data1$Price) - min(data1$Price)) / data.half
  hl2 <- (max(data2$Price) - min(data2$Price)) / data.half
  hl <- (max(data$Price) - min(data$Price)) / nrow(data)
  
  D <- (log(hl1 + hl2) - log(hl)) / log(2)
  D
}

#Get for all ETFs the fractal dimension 'D'
fractal_dimensions <- sapply(etfs.red, function(x) fractal_dimension(x))

#First look at fractal dimensions
  #table(round(fractal_dimensions, 1))
```

The fractal dimension can be thought of a measure of volatility for a finencial price chart. A more common measure of volatility is the *Historic Volatility*. 
Historic Volatility ($=HV$) has been calculated as:
$$\begin{aligned}
HV &= sd(R) \\
R &= ln(\frac{V_f}{V_i})
\end{aligned}$$

$R =$ logarithmic return  
$V_i =$ price when market closed on day i  
$V_f =$ price when market closed of the next day  


```{r Historic Volatility}
#Logarithmic or continuously compounded return
get_log_return <- function(x){
  nextP <- c(0, x$Price[1:length(x$Price)-1])
  x[, log_return := log(nextP/Price)]
}

get_abs_return <- function(x){
  nextP <- c(0, x$Price[1:length(x$Price)-1])
  x[, abs_return := nextP-Price]
}

etfs.red <- lapply(etfs.red, get_log_return)
etfs.red <- lapply(etfs.red, get_abs_return)

#Historic Volatility
hist_vol <- sapply(etfs.red, function(x) sd(x$log_return[-1]))
```

```{r Read in S & P 500}
sp500 <- fread("sp500.csv", sep = ",")
sp500[, Price := apply(sp500, 1, function(x) mean(as.numeric(c(x["High"], x["Low"]))))]
sp500 <- sp500[, c("Date", "Price")]
sp500$Date <- as.Date(sp500$Date)

get_log_return(sp500)
sp500$log_return[1] <- 0
```

```{r Get betas}
sp500.var <- var(sp500$log_return[-1])

get_beta <- function(x, sp500.var){
  x$log_return[1] <- 0
  #Merge to get overlaping Dates
  a <- merge(x, sp500, by = "Date")
  x.cov <- cov(a$log_return.x, a$log_return.y)
  beta <- x.cov/sp500.var
  beta
}

betas <- sapply(etfs.red, function(x) get_beta(x, sp500.var))
```


```{r sharpe ratio}

#Calculating sharpe ration S = mean(daily_return - daily_rf) / std(daily_return) 
#daily_rf = 0%

calc_sharpe_ratio <- function(x, abs = FALSE){
  if(abs == FALSE){
 sharpe <- mean(x$log_return[-1])/sd(x$log_return[-1])
 sharpe
} else {
    sharpe <- mean(x$abs_return[-1])/sd(x$abs_return[-1])
    sharpe
  }
}

sharpe_ratios <- sapply(etfs.red, function(x) calc_sharpe_ratio(x))
sharpe_ratios_abs <- sapply(etfs.red, function(x) calc_sharpe_ratio(x, abs = TRUE))

```

```{r Volatility measures Histogram, warning=FALSE}
volatility_measures <- data.table(etf = names(etfs.red), hist_vol, fractal_dimensions, betas, sharpe_ratios_abs, overall_returns)

volatility_measures_hist <- melt(volatility_measures[, -c("etf", "overall_returns")], variable.name = "measure")
levels(volatility_measures_hist$measure) <- c("Historic Volatilitys", "Fractal Dimensions", "Betas", "Sharpe Ratios")
ggplot(volatility_measures_hist, aes(value)) + 
  geom_histogram(bins = 20, col = "black", fill = "lightgrey") + 
  facet_wrap(~measure, scales = "free") +
  labs(title = "Distrubution of volatility measurments") + 
  theme_minimal()
```


In the following plot I compare the ETFs with the lowest and highest fractal dimensions. Additionally to the Price chart I plotted the daily price fluctuation. This is simply the change of the Price from one day to the next. Note that the scales for the price charts are different!

```{r calc price change}
#Highest vs lowest fractal dimension
highest.dimension <- fractal_dimensions[which.max(fractal_dimensions)]
lowest.dimension <- fractal_dimensions[which.min(fractal_dimensions)]

highest.dimension.chart <- etfs.red[[attributes(highest.dimension)$names]]
lowest.dimension.chart <- etfs.red[[attributes(lowest.dimension)$names]]

price_fluc <- function(price.chart){
  next.price <- c(NA, price.chart[1:nrow(price.chart)-1]$Price)
  price.chart[, price_change := Price - next.price]
}

price_fluc(lowest.dimension.chart)
price_fluc(highest.dimension.chart)

```

```{r lowest D, warning=FALSE}
p <- lowest.dimension.chart %>% 
  plot_ly(x = ~Date, y = ~Price) %>% add_lines()

p1 <- lowest.dimension.chart %>% 
  plot_ly(x = ~Date, y = ~price_change, type = "bar") %>%
  layout(yaxis = list(title = "Price Change"))

subplot(p, p1, nrows = 2, shareX = T, titleY = T) %>% 
  layout(title=  paste("Price chart of rwk.us (D = ", round(lowest.dimension, 3),")", sep = ""), showlegend = F)
```

```{r highest D, warning=FALSE}
p <- highest.dimension.chart %>% 
  plot_ly(x = ~Date, y = ~Price) %>% add_lines()

p1 <- highest.dimension.chart %>% 
  plot_ly(x = ~Date, y = ~price_change, type = "bar") %>%
  layout(yaxis = list(title = "Price Change"))

subplot(p, p1, nrows = 2, shareX = T, titleY = T) %>% 
  layout(title= paste("Price chart of emlp.us (D = ", round(highest.dimension, 3), ")", sep = ""), showlegend = F)
```


```{r HV}
highest.hv.chart <- etfs.red[[which.max(hist_vol)]]
lowest.hv.chart <- etfs.red[[which.min(hist_vol)]]

highest_hv <- hist_vol[which.max(hist_vol)]
lowest_hv <- variabl[which.min(hist_vol)]

price_fluc(highest.hv.chart)
price_fluc(lowest.hv.chart)
```


```{r highest HV, warning=FALSE}
p <- highest.hv.chart %>% 
  plot_ly(x = ~Date, y = ~Price) %>% add_lines()

p1 <- highest.hv.chart %>% 
  plot_ly(x = ~Date, y = ~price_change, type = "bar") %>%
  layout(yaxis = list(title = "Price Change"))

subplot(p, p1, nrows = 2, shareX = T, titleY = T) %>% 
  layout(title=paste("Price chart of ", names(highest_hv) ," (HV = ", round(highest_hv, 4), ")", sep = ""), showlegend = F)
```

```{r lowest HV, warning=FALSE}
p <- lowest.hv.chart %>% 
  plot_ly(x = ~Date, y = ~Price) %>% add_lines()

p1 <- lowest.hv.chart %>% 
  plot_ly(x = ~Date, y = ~price_change, type = "bar") %>%
  layout(yaxis = list(title = "Price Change"))

subplot(p, p1, nrows = 2, shareX = T, titleY = T) %>% 
  layout(title=paste("Price chart of ", names(lowest_hv) ," (HV = ", round(lowest_hv, 4), ")", sep = ""), showlegend = F)
```

```{r price chart plot}
plot_price_chart <- function(etfs, variable, max = TRUE){
  require(magrittr)
  require(plotly)
  require(data.table)
  
  if(max == TRUE){
  
    max.chart <- etfs[[which.max(variable)]]
    max_var <- variable[which.max(variable)]
    max.chart <- price_fluc(max.chart)
    
    p <- max.chart %>% plot_ly(x = ~Date, y = ~Price) %>% add_lines()

    p1 <- max.chart %>% 
      plot_ly(x = ~Date, y = ~price_change, type = "bar") %>%
      layout(yaxis = list(title = "Price Change"))

    subplot(p, p1, nrows = 2, shareX = T, titleY = T) %>% 
      layout(title=paste("Price chart of ", names(max_var) ," (HV = ", round(max_var, 4), ")", sep = ""), showlegend = F)

  } else {
    
    min.chart <- etfs[[which.min(variable)]]
    min_var <- variable[which.min(variable)]
    min.chart <- price_fluc(min.chart)
    
    p <- min.chart %>% plot_ly(x = ~Date, y = ~Price) %>% add_lines()

    p1 <- min.chart %>% 
      plot_ly(x = ~Date, y = ~price_change, type = "bar") %>%
      layout(yaxis = list(title = "Price Change"))

    subplot(p, p1, nrows = 2, shareX = T, titleY = T) %>% 
      layout(title=paste("Price chart of ", names(min_var) ," (HV = ", round(min_var, 4), ")", sep = ""), showlegend = F)
    
  }
  

}
  
plot_price_chart(etfs.red, hist_vol, max = T)
plot_price_chart(etfs.red, hist_vol, max = F)
plot_price_chart(etfs.red, fractal_dimensions, max = T)
plot_price_chart(etfs.red, fractal_dimensions, max = F)
plot_price_chart(etfs.red, betas, max = T)
plot_price_chart(etfs.red, betas, max = F)
plot_price_chart(etfs.red, sharpe_ratios_abs, max = T)
plot_price_chart(etfs.red, sharpe_ratios_abs, max = F)

```



```{r}

pairs.panels(volatility_measures[, -c("etf")])

#lapply(a$etf, function(x) plot( etfs.red[[ x ]]$Price , type = "l"))
```


```{r}

g_legend <- function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)}

legend <- ggplot(volatility_measures, 
       aes(hist_vol, betas, color = overall_returns)) + 
  geom_point() +
  scale_color_continuous(low = "lightgrey", high = "black")


legend <- g_legend(legend)

p1 <- ggplot(volatility_measures, 
       aes(hist_vol, betas, size = overall_returns, color = overall_returns)) + 
  geom_point(aes(alpha = 0.3), show.legend = F) +
  scale_color_continuous(low = "lightgrey", high = "black") + 
  theme_minimal()


p2 <- ggplot(volatility_measures,
        aes(hist_vol, sharpe_ratios_abs, size = overall_returns, color = overall_returns)) +
  geom_point(aes(alpha = 0.3), show.legend = F) +
  scale_color_continuous(low = "lightgrey", high = "black") + 
  theme_minimal()

p3 <- ggplot(volatility_measures,
       aes(hist_vol, fractal_dimensions, size = overall_returns, color = overall_returns)) +
  geom_point(aes(alpha = 0.3), show.legend = F) +
  scale_color_continuous(low = "lightgrey", high = "black") + 
  theme_minimal()


p4 <- ggplot(volatility_measures,
       aes(betas, sharpe_ratios_abs, size = overall_returns, color = overall_returns)) +
  geom_point(aes(alpha = 0.3), show.legend = F) +
  scale_color_continuous(low = "lightgrey", high = "black") + 
  theme_minimal()

p5 <- ggplot(volatility_measures,
       aes(betas, fractal_dimensions, size = overall_returns, color = overall_returns)) +
  geom_point(aes(alpha = 0.3), show.legend = F) +
  scale_color_continuous(low = "lightgrey", high = "black") + 
  theme_minimal()

p6 <- ggplot(volatility_measures,
       aes(sharpe_ratios_abs, fractal_dimensions, size = overall_returns, color = overall_returns)) +
  geom_point(aes(alpha = 0.3), show.legend = F) +
  scale_color_continuous(low = "lightgrey", high = "black") +
  theme_minimal()

layout <- rbind(c(1,NA,7),c(2,4,NA),c(3,5,6))

grid.arrange(p1, p2, p3, p4, p5, p6, legend, layout_matrix = layout)



```

```{r, include=FALSE}

merged_dt <- etfs.red[[1]]
merged_dt <- merged_dt[, c("Date", "Price")]

first.name <- names(etfs.red[1])
colnames(merged_dt) <- c("Date", first.name)


#Tales Long. only one date in common for all etfs with > 1000 obs
for(etf in 1:length(etfs.red)){
  x <- etfs.red[[etf]]
  x <- x[, c("Date", "Price")]
  x.name <- names(etfs.red[etf])
  colnames(x) <- c("Date", x.name)
  merged_dt <- merge(merged_dt, x, by = "Date")  
}


top_D <- sort(fractal_dimensions)[1:10]
top_D_dt <- etfs.red[[names(top_D[1])]]
top_D_dt <- top_D_dt[, c("Date", "Price")]

first.name <- names(top_D[1])
colnames(top_D_dt) <- c("Date", first.name)

for(etf in 2:length(top_D)){
  x <- etfs.red[[names(top_D[etf])]]
  x <- x[, c("Date", "Price")]
  x.name <- names(top_D[etf])
  colnames(x) <- c("Date", x.name)
  top_D_dt <- merge(top_D_dt, x, by = "Date")
}


top_D_mat <- as.matrix(top_D_dt[, -c("Date")])

top_D_cor <- cor(top_D_mat)

corrplot(top_D_cor)

```

